---
title: "Final (outline)"
author: "Amanda Hernandez, Ruby Hickman, Abby Zhong"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	fig.width = 10,
	message = FALSE,
	warning = FALSE
)

pacman::p_load(tidyverse, mgcv, splines, broom, broom.mixed, ggeffects, gtsummary, 
               flextable, cowplot)


#set some global preferences
theme_set(theme_bw())
options(digits = 4)


tbl_theme <- function(dat, gt = FALSE, caption = "", footer = ""){
  if(gt == FALSE) {
    dat_flex <- dat %>% 
    flextable()
  }
  else {
    dat_flex <- dat %>% 
    as_flex_table(show_coltype = FALSE) 
  }
  
  dat_flex %>% 
    align(align = "left", part = "all") %>% 
    set_caption(caption = caption) %>% 
    add_footer_lines(footer) %>% 
    color(part = "footer", color = "#666666") %>% 
    set_table_properties(layout = "autofit", width = 1)
}

cat_columns <- c("WARMCOLD", "WKDAY", "RACE", "TWODRINK", 
                 "SMK", "CHD", "DIABETE", "FBG", "STATIN")

nas_bp <- readRDS("nas_bp.rds") %>%
  mutate(across(any_of(cat_columns), factor))




```



- What kind of models are appropriate for your study question and dataset? Be sure to include how you specified your model (e.g. degrees of freedom, distribution family, etc.) and any transformations you performed. Note that you can use `SINETIME+COSTIME+as.factor(WKDAY)` simultaneously in your model to control for seasonality.

```{r explore}

skimr::skim(nas_bp)

tbl_summary(nas_bp, 
            include = -c("ID", "DATE", "YEAR", "SINETIME", "COSTIME")) %>% 
  tbl_theme(gt = TRUE) 


nas_bp %>% 
  select(-any_of(cat_columns)) %>% 
  pivot_longer(names_to = "variable", values_to = "value", -ID) %>% 
  subset(variable %in% c("SBP", "TEMPC24H", "BC24H")) %>% 
           ggplot(aes(x = value)) + 
           geom_histogram(aes(y = after_stat(density)),
                          color = "black",
                          fill = "white",
                          bins = 10) +
           geom_density(fill = "red", alpha = 0.25)+ 
           facet_wrap(~variable, scales = "free")


```




- Explore the relationship of each continuous covariate with SBP. Show them when appropriate to justify your model specification (e.g., are they linear? Non-linear?)

```{r mod1}

# model 1 -- exposure + confounders + predictors w splines

mod1_temp <- gamm(SBP ~ s(TEMPC24H, bs = 'cr', fx = FALSE) + 
                    s(RHUM24H, bs = 'cr', fx = FALSE) + 
                    s(BMI, bs = 'cr', fx = FALSE) + 
                    s(AGE, bs = 'cr', fx = FALSE) + 
                    s(NEDUC, bs = 'cr', fx = FALSE) + 
                    SMK + TWODRINK + STATIN + FBG + 
                    # control for seasonality 
                    SINETIME + COSTIME + WKDAY, 
                  random = list(ID =  ~ 1),
                  family = gaussian(), data = nas_bp)


as_flextable(mod1_temp$gam)
# all edfs = 1, so they are linear and we can remove the splines

mod1_bc <- gamm(SBP ~ BC24H + TEMPC24H + 
                  RHUM24H + 
                  BMI + 
                  s(AGE, bs = 'cr', fx = FALSE) + 
                  NEDUC + 
                  SMK + TWODRINK + STATIN + FBG + 
                  # control for seasonality 
                  SINETIME + COSTIME + WKDAY, 
                random = list(ID =  ~ 1),
                family = gaussian(), data = nas_bp)


as_flextable(mod1_bc$gam)

# all edfs = 1 except for age, so they are linear and we can remove the splines

```

```{r mod2}

# model 2 -- exp + confounders + precision vars wo splines

mod2_covars <- c(" +  RHUM24H + BMI + NEDUC + s(AGE, bs = 'cr', fx = FALSE) + SMK + TWODRINK + STATIN + FBG + SINETIME + COSTIME + WKDAY")

mod2_temp <- gamm(as.formula(paste0("SBP ~ TEMPC24H", mod2_covars)),
                       random = list(ID =  ~ 1),
                       family = gaussian(),
                       data = nas_bp)


as_flextable(mod2_temp$gam)

plot(mod2_temp$gam)


mod2_bc <- gamm(as.formula(paste0("SBP ~ BC24H + TEMPC24H", mod2_covars)),
                       random = list(ID =  ~ 1),
                       family = gaussian(),
                       data = nas_bp)


as_flextable(mod2_bc$gam)

plot(mod2_bc$gam)


```



- Present the dose-response relationship of same-day BC and temperature with SBP. (Is it linear?)

```{r DR}


dr_mod2_temp <- ggpredict(mod2_temp$gam, terms = "TEMPC24H") %>% 
  plot()

dr_mod2_bc <- ggpredict(mod2_bc$gam, terms = "BC24H") %>% 
  plot()

plot_grid(dr_mod2_temp, dr_mod2_bc)

```


- Compute an effect estimate (and 95% CI) for the same-day BC and temperature with SBP. 

```{r est}

eff_estimate <- function(mod, term_keep){
  broom.mixed::tidy(mod, parametric = TRUE, conf.int = TRUE) %>% 
  filter(term %in% c(term_keep)) %>% 
  tbl_theme()
}

eff_estimate(mod2_temp$gam, term_keep = "TEMPC24H")

eff_estimate(mod2_bc$gam, term_keep = "BC24H")


```


- Does diabetes status modify the effect of same-day BC exposure on SBP? If so, provide the effect estimates (with 95% CI) between BC and SBP among diabetics and non-diabetics, respectively.

```{r diab_emm}




mod2_temp_diab <- gamm(as.formula(paste0("SBP ~ TEMPC24H*DIABETE", mod2_covars)),
                       random = list(ID =  ~ 1),
                       family = gaussian(),
                       data = nas_bp)


mod2_bc_diab <- gamm(as.formula(paste0("SBP ~ BC24H*DIABETE + TEMPC24H", mod2_covars)),
                       random = list(ID =  ~ 1),
                       family = gaussian(),
                       data = nas_bp)


as_flextable(mod2_temp_diab$gam)
as_flextable(mod2_bc_diab$gam)


#it's not significant, so we probs don't have to report est and CIs -- so I didn't update this code, but I can if we want them
#
# get estimates and std errors (if we decide to report)
# mod2_diab_interaction_stderrs <- data.frame(mod2_diab$lme$varFix) %>% 
#   rownames_to_column("term") %>% 
#   pivot_longer(names_to = "term2", values_to = "stderr", -term) %>% 
#   filter(str_detect(term, "BC24H") & str_detect(term2, "BC24H"))
# 
# broom.mixed::tidy(mod2_diab$gam, parametric = TRUE, conf.int = TRUE) %>% 
#   filter(str_detect(term, "BC24H")) %>% 
#   select(term, estimate) %>% 
#   group_by(.) %>% 
#   mutate(estimate = case_when(term == "BC24H:DIABETE1" ~ sum(estimate),
#                               TRUE ~ estimate)) %>% 
#   left_join(data.frame(term = c("BC24H", "BC24H:DIABETE1"),
#                        std.err = c(sqrt(mod2_diab_interaction_stderrs$stderr[which(mod2_diab_interaction_stderrs$term == "XBC24H" &
#                                                                                      mod2_diab_interaction_stderrs$term2 == "XBC24H")]),
#                                    sqrt(sum(mod2_diab_interaction_stderrs$stderr))))) %>% 
#   tbl_theme()
# 
# 


```


- Does the association between SBP and same-day temperature vary by warm and cold seasons? How does the dose-response relationship differ between warm and cold seasons? 

```{r warmcold}

# effect modification by WARMCOLD

mod2_temp_warmcold <- gamm(as.formula(paste0("SBP ~ TEMPC24H*WARMCOLD", mod2_covars)),
                       random = list(ID =  ~ 1),
                       family = gaussian(),
                       data = nas_bp)

as_flextable(mod2_temp_warmcold$gam)

warmcold0_temp <- ggpredict(mod2_temp_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = "0"))
warmcold1_temp <- ggpredict(mod2_temp_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = "1"))


ggplot(warmcold0_temp) +
  geom_line(aes(x, y = predicted, color = "Cold season")) +
  geom_ribbon(aes(x = x, ymin = conf.low, ymax = conf.high), fill = "blue", alpha = .1) + 
  geom_line(warmcold1_temp, mapping = aes(x, y = predicted, color = "Warm season")) + 
  geom_ribbon(warmcold1_temp, mapping = aes(x = x, ymin = conf.low, ymax = conf.high), fill = "red", alpha = .1) + 
  scale_color_manual(name = "Season", values = c("Cold season" = "blue", "Warm season" = "red")) + 
  labs(x = "TEMPC24H", y = "Predicted SBP")


```


- Are there any subgroups that are particularly vulnerable to the effect of same-day BC exposure on SBP?

```{r emm}

model_fun <- list()
mod2_bc_emm <- list()
int_term <- list()
for(i in cat_columns){
  
  model_fun[[i]] <- paste0("SBP ~ TEMPC24H + BC24H + ", i, "*BC24H", mod2_covars)
  
  mod2_bc_emm[[i]] <- gamm(as.formula(model_fun[[i]]),
                    random = list(ID =  ~ 1),
                    family = gaussian(), data = nas_bp)
  
  int_term[[i]] <- tidy(mod2_bc_emm[[i]]$gam, parametric = TRUE) %>% 
    filter(str_detect(term, ":"))
  
}

#inspect interaction terms significance
bind_rows(int_term, .id = "model") %>% 
  mutate(p = case_when(p.value < 0.001 ~ "***",
                       p.value < 0.01 ~ "**",
                       p.value < 0.05 ~ "*",
                              TRUE ~ ""))


```


Are there lagged effects of temperature and BC exposure on SBP?

```{r lag}

mod2_temp_lag <- gamm(as.formula(paste0("SBP ~ TEMPC24H + TMPCD1 + TMPCD2", mod2_covars)),
                 random = list(ID =  ~ 1),
                 family = gaussian(), data = nas_bp)

mod2_bc_lag <- gamm(as.formula(paste0("SBP ~ TEMPC24H + BC24H + BCD1 + BCD2", mod2_covars)),
                 random = list(ID =  ~ 1),
                 family = gaussian(), data = nas_bp)

as_flextable(mod2_temp_lag$gam)
as_flextable(mod2_bc_lag$gam)
# we can just interpret these estimates and CIs and p-values as they are!


mod2_temp_MA<- gamm(as.formula(paste0("SBP ~ TMPCMA2", mod2_covars)),
                 random = list(ID =  ~ 1),
                 family = gaussian(), data = nas_bp)


mod2_bc_MA<- gamm(as.formula(paste0("SBP ~ TEMPC24H + BCMA2", mod2_covars)),
                 random = list(ID =  ~ 1),
                 family = gaussian(), data = nas_bp)

as_flextable(mod2_temp_MA$gam)
as_flextable(mod2_bc_MA$gam)
# assume that the effect is the same across the two days (if you use the moving avg)



```

