---
title: "Final (outline)"
author: "Amanda Hernandez, Ruby Hickman, Abby Zhong"
date: "2023-03-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	fig.width = 10,
	message = FALSE,
	warning = FALSE
)

pacman::p_load(tidyverse, mgcv, splines, broom, ggeffects, gtsummary, 
               flextable, cowplot)


#set some global preferences
theme_set(theme_bw())
options(digits = 4)
tbl_theme <- function(dat, gt = FALSE, caption = "", footer = ""){
  if(gt == FALSE) {
    dat_flex <- dat %>% 
    flextable()
  }
  else {
    dat_flex <- dat %>% 
    as_flex_table(show_coltype = FALSE) 
  }
  
  dat_flex %>% 
    align(align = "left", part = "all") %>% 
    set_caption(caption = caption) %>% 
    add_footer_lines(footer) %>% 
    color(part = "footer", color = "#666666") %>% 
    set_table_properties(layout = "autofit", width = 1)
}

cat_columns <- c("WARMCOLD", "WKDAY", "RACE", "TWODRINK", 
                 "SMK", "CHD", "DIABETE", "FBG", "STATIN")

nas_bp <- readRDS("nas_bp.rds") %>% 
  mutate(across(any_of(cat_columns), factor))



```



- What kind of models are appropriate for your study question and dataset? Be sure to include how you specified your model (e.g. degrees of freedom, distribution family, etc.) and any transformations you performed. Note that you can use `SINETIME+COSTIME+as.factor(WKDAY)` simultaneously in your model to control for seasonality.

```{r explore}

skimr::skim(nas_bp)

tbl_summary(nas_bp, 
            include = -c("ID", "DATE", "YEAR", "SINETIME", "COSTIME")) %>% 
  tbl_theme(gt = TRUE) 


nas_bp %>% 
  select(-any_of(cat_columns)) %>% 
  pivot_longer(names_to = "variable", values_to = "value", -ID) %>% 
  subset(variable %in% c("SBP", "TEMPC24H", "BC24H")) %>% 
           ggplot(aes(x = value)) + 
           geom_histogram(aes(y = after_stat(density)),
                          color = "black",
                          fill = "white",
                          bins = 30) +
           geom_density(fill = "red", alpha = 0.25)+ 
           facet_wrap(~variable, scales = "free")
 
# log transform SBP         
ggplot(nas_bp, aes(x = log(SBP))) +
  geom_histogram(aes(y = after_stat(density)),
                 color = "black",
                 fill = "white",
                 bins = 30) +
  geom_density(fill = "red", alpha = 0.25)



```

```{r mod1}

# model 1 -- exposure + confounders only

mod1 <- gamm(log(SBP) ~ s(TEMPC24H, bs = 'cr', fx = FALSE) + 
               s(BC24H, bs = 'cr', fx = FALSE) + 
               s(RHUM24H, bs = 'cr', fx = FALSE) + 
               # control for seasonality 
               SINETIME + COSTIME + WKDAY, 
             random = list(ID =  ~ 1),
             family = gaussian(), data = nas_bp)


as_flextable(mod1$gam)
# all edfs = 1, so they are linear and we can remove the splines

```


- Explore the relationship of each continuous covariate with SBP. Show them when appropriate to justify your model specification (e.g., are they linear? Non-linear?)


```{r mod2}

# model 2 -- exp + confounders + predictors 

# let's examine non-linearity with continuous variable predictors and SBP 
#remember to make a spline for TEMPC24H and BC24H 

mod2 <- gamm(log(SBP) ~ TEMPC24H + BC24H + RHUM24H + 
               s(AGE, bs = 'cr', fx = FALSE) + 
               RACE + TWODRINK + CHD + DIABETE +
               s(BMI, bs = 'cr', fx = FALSE) + SMK + FBG + 
               s(NEDUC, bs = 'cr', fx = FALSE) + STATIN + 
               # control for seasonality 
               SINETIME + COSTIME + WKDAY, 
             random = list(ID =  ~ 1),
             family = gaussian(), data = nas_bp)

#should we be including all the predictors? should we justify our decision based 
#on p-values? 

#get edfs
as_flextable(mod2$gam)

```

```{r mod3}

# model 3 -- exp + confounders + predictors (wo splines for linear terms) 

# we don't need a spline for BMI or NEDUC -- its linear -- but we do for age

mod3 <- gamm(log(SBP) ~ TEMPC24H + BC24H + RHUM24H + 
               s(AGE, bs = 'cr', fx = FALSE) + 
               RACE + TWODRINK + CHD + DIABETE +
               BMI + SMK + FBG + 
               NEDUC + STATIN + 
               # control for seasonality 
               SINETIME + COSTIME + WKDAY,
             random = list(ID =  ~ 1),
             family = gaussian(), data = nas_bp)

as_flextable(mod3$gam)

```



- Present the dose-response relationship of same-day BC and temperature with SBP. (Is it linear?)

```{r DR}

ggpredict(mod3$gam, terms = "TEMPC24H") %>% 
  plot()

ggpredict(mod3$gam, terms = "BC24H") %>% 
  plot()

```


- Compute an effect estimate (and 95% CI) for the same-day BC and temperature with SBP. 

```{r est}

tidy(mod3$gam, parametric = TRUE, conf.int = TRUE) %>% 
  filter(term %in% c("TEMPC24H", "BC24H")) %>% 
  tbl_theme()

```


- Does diabetes status modify the effect of same-day BC exposure on SBP? If so, provide the effect estimates (with 95% CI) between BC and SBP among diabetics and non-diabetics, respectively.

```{r diab_emm}

mod3_diab <- gamm(log(SBP) ~ TEMPC24H + BC24H + RHUM24H + 
                    s(AGE, bs = 'cr', fx = FALSE) + 
                    RACE + TWODRINK + CHD + DIABETE +
                    BMI + SMK + FBG + 
                    NEDUC + STATIN + 
                    DIABETE*BC24H + 
                    # control for seasonality 
                    SINETIME + COSTIME + WKDAY,
                  random = list(ID =  ~ 1),
                  family = gaussian(), data = nas_bp)

# get estimates and std errors (if we decide to report)
mod3_diab_interaction_stderrs <- data.frame(mod3_diab$lme$varFix) %>% 
  rownames_to_column("term") %>% 
  pivot_longer(names_to = "term2", values_to = "stderr", -term) %>% 
  filter(str_detect(term, "BC24H") & str_detect(term2, "BC24H"))

broom.mixed::tidy(mod3_diab$gam, parametric = TRUE, conf.int = TRUE) %>% 
  filter(term %in% c("BC24H", "BC24H:DIABETE")) %>% 
  select(term, estimate) %>% 
  group_by(.) %>% 
  mutate(estimate = case_when(term == "BC24H:DIABETE" ~ sum(estimate),
                              TRUE ~ estimate)) %>% 
  left_join(data.frame(term = c("BC24H", "BC24H:DIABETE"),
                       std.err = c(sqrt(mod3_diab_interaction_stderrs$stderr[which(mod3_diab_interaction_stderrs$term == "XBC24H" & mod3_diab_interaction_stderrs$term2 == "XBC24H")]),
                                   sqrt(sum(mod3_diab_interaction_stderrs$stderr))))) %>% 
  tbl_theme()




```


- Does the association between SBP and same-day temperature vary by warm and cold seasons? How does the dose-response relationship differ between warm and cold seasons? 

```{r warmcold}

# is this asking about effect modification by WARMCOLD? 

mod3_warmcold <- gamm(log(SBP) ~ TEMPC24H + BC24H + RHUM24H + 
                    s(AGE, bs = 'cr', fx = FALSE) + 
                    RACE + TWODRINK + CHD + DIABETE +
                    BMI + SMK + FBG + 
                    NEDUC + STATIN + WARMCOLD + 
                    WARMCOLD*TEMPC24H + 
                    # control for seasonality 
                    SINETIME + COSTIME + WKDAY,
                  random = list(ID =  ~ 1),
                  family = gaussian(), data = nas_bp)

as_flextable(mod3_warmcold$gam)

warmcold0 <- ggpredict(mod3_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = 0))
warmcold1 <- ggpredict(mod3_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = 1))

ggplot(warmcold0) +
  #geom_point(nas_bp, mapping = aes(x = TEMPC24H, y = SBP), alpha = 0.1) +
  geom_line(aes(x, y = predicted, color = "Cold season")) +
  geom_ribbon(aes(x = x, ymin = conf.low, ymax = conf.high), fill = "blue", alpha = .1) + 
  geom_line(warmcold1, mapping = aes(x, y = predicted, color = "Warm season")) + 
  geom_ribbon(warmcold1, mapping = aes(x = x, ymin = conf.low, ymax = conf.high), fill = "red", alpha = .1) + 
  scale_color_manual(name = "Season", values = c("Cold season" = "blue", "Warm season" = "red"))

# can also show them side by side 
# warmcold0 <- ggpredict(mod3_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = 0)) %>% 
#   plot() + 
#   ylim(125, 150) + 
#   ggtitle("Predicted values of SBP (Cold season)")
# 
# warmcold1 <- ggpredict(mod3_warmcold$gam, terms = "TEMPC24H", condition = c(WARMCOLD = 1)) %>% 
#   plot() + 
#   ylim(125, 150) + 
#   ggtitle("Predicted values of SBP (Warm season)")
# 
# 
# plot_grid(warmcold0, warmcold1)


```


- Are there any subgroups that are particularly vulnerable to the effect of same-day BC exposure on SBP?


```{r emm}

# I'm assumming we want to test for EMM for subgroups -- do this for all categorical subgroups? 

model_fun <- list()
mod3_emm <- list()
int_term <- list()
for(i in cat_columns){
  
  model_fun[[i]] <- paste0("log(SBP) ~ TEMPC24H + BC24H + RHUM24H + s(AGE, bs = 'cr', fx = FALSE) + RACE + TWODRINK + CHD + DIABETE + BMI + SMK + FBG + NEDUC + STATIN + ",
                      i, "*BC24H + SINETIME + COSTIME + WKDAY")
  
  mod3_emm[[i]] <- gamm(as.formula(model_fun[[i]]),
                    random = list(ID =  ~ 1),
                    family = gaussian(), data = nas_bp)
  
  int_term[[i]] <- tidy(mod3_emm[[i]]$gam, parametric = TRUE) %>% 
    filter(str_detect(term, ":"))
  
}

#inspect interaction terms significance
bind_rows(int_term, .id = "model") %>% 
  mutate(p = case_when(p.value < 0.001 ~ "***",
                       p.value < 0.01 ~ "**",
                       p.value < 0.05 ~ "*",
                              TRUE ~ ""))


```


Are there lagged effects of temperature and BC exposure on SBP?

```{r lag}
# ¯\ (ツ) /¯
mod3_lag <- gamm(log(SBP) ~ TEMPC24H + TMPCD1 + TMPCD2 + BC24H + BCD1 + BCD2 +
                   RHUM24H + s(AGE, bs = 'cr', fx = FALSE) + 
                   RACE + TWODRINK + CHD + DIABETE + BMI + SMK + FBG + 
                   NEDUC + STATIN + 
                   # control for seasonality 
                   SINETIME + COSTIME + WKDAY,
                 random = list(ID =  ~ 1),
                 family = gaussian(), data = nas_bp)

as_flextable(mod3_lag$gam)

```

